---
title: "Reproducible Research: Peer Assessment 1"
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
output: 
  html_document:
    keep_md: true
---

```{r library, echo=TRUE}
library(ggplot2)
```

## Loading and preprocessing the data
Load the data from the activity.csv file.
```{r loadingdata, echo=TRUE}
act <- read.csv( unz("activity.zip", "activity.csv") )
complete <- sum(complete.cases(act) == TRUE)    # number of obs complete (no NA's)
missing <- sum(!complete.cases(act) == TRUE)    # number of obs missing values (with NA's)
str(act)
```
Total observations in the data with **completed cases** is `r complete` and **missing cases** (NA) is `r missing`.



## What is mean total number of steps taken per day?
#### 1. Calculate the total number of steps taken per day
```{r totalsteps, echo=TRUE}
steps.sum.date <- tapply(act$steps, act$date, sum)  # sum of steps over subsetting by each date
avg <- mean(steps.sum.date, na.rm=TRUE)             # mean total # of steps taken each day
med <- median(steps.sum.date, na.rm=TRUE)           # median total # of steps taken each day   
df <- data.frame(totalsteps=steps.sum.date, date=names(steps.sum.date))
summary(df$totalsteps)
```

#### 2. Make a histogram of the total number of steps taken each day
```{r histogramtotalsteps, echo=TRUE, fig.width=12}
g <- ggplot(data=df, aes(x=totalsteps))
g + geom_histogram(binwidth=550, col="black", fill="grey") +
    geom_vline(xintercept=avg, size=0.5, col="red") +
    geom_vline(xintercept=med, size=0.5, col="blue") +
    labs(title = "Histogram of Total Steps taken per Day",
         x = "Number of Daily Steps")
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r meantotalsteps, echo=TRUE}
avg <- round( mean(df$totalsteps, na.rm=TRUE) )
med <- median(df$totalsteps, na.rm=TRUE)
print(avg); print(med)
```
The **mean** value is $`r round(avg, 0)`$ and the **median** value is $`r round(med, 0)`$.  
*Note: In the histogram above, the mean is denoted by the red color vertical line and the median by the blue color vertical line.*



## What is the average daily activity pattern?
#### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r timeseries, echo=TRUE, fig.width=10}
split.interval <- split(act, act$interval)                           # split data by interval
interval.avgsteps <- sapply( split.interval, 
                             function(x) mean(x$steps, na.rm=TRUE) ) # get average steps for each interval
df <- data.frame(interval=as.numeric( names(interval.avgsteps) ), 
                 avgsteps=interval.avgsteps)
idxmax <- which.max(df$avgsteps)                        # row index of df where this max exists
intervalmax <- df[idxmax, ]$interval                    # interval (x-axis) for max avg # steps
g <- ggplot(data=df, aes(x=interval, y=avgsteps))
g + geom_line() +
    geom_vline(xintercept=intervalmax, size=0.5, col="blue") +
    labs(title = "Time Series of 5-Minute Interval and Average Steps",
         x = "5-min Interval",
         y = "Average Number of Steps")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r maxsteps, echo=TRUE}
stepsmax <- max(df$avgsteps)                # max avg # of steps (y-axis)
idxmax <- which.max(df$avgsteps)            # row index of df where this max exists
intervalmax <- df[idxmax, ]$interval        # interval (x-axis) for max avg # of steps
print(stepsmax); print(intervalmax)
```
The **maximum average number of steps** (y-axis) is $`r round(stepsmax, 0)`$.  
The **5-min interval** (x-axis) for this maximum value is `r intervalmax`.  
*Note: In the time series plot above, the maximum value and interval is denoted by the blue color vertical line.*



## Imputing missing values
#### 1. Calculate and report the total number of missing values in the dataset.
```{r missingvalues, echo=TRUE}
missing <- sum(!complete.cases(act))        # number of observations missing values (with NA's)
```
The total number of rows in **original dataset** with missing values (NAs) is `r missing`.

#### 2. Devise a strategy for filling in all of the missing values in the dataset.  
We will use the mean for that 5-minute interval to replace all occurances of NA in the steps variable for the same interval value. Eg if interval 2245 has 8 NA's, then we replace all 8 with the average number of steps at interval 2245.  

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r imputedataset, echo=TRUE}
split.interval <- split(act, act$interval)  
for(i in 1:nrow(df)) {
    split.interval[i][[1]]$steps[is.na(split.interval[i][[1]]$steps)] <- round( df[i,]$avgsteps )
}
act.imp <- unsplit(split.interval, act$interval)    # new dataset (imputed)
missingi <- sum(!complete.cases(act.imp))         # total # of NA's in new imputed dataset (should be 0)
```
The total number of rows in **new dataset** with missing values (NAs) is `r missingi`.

#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r newtotalsteps, echo=TRUE}
steps.sum.date.imp <- tapply(act.imp$steps, act.imp$date, sum)  # sum of steps over subsetting by each date
avgi <- mean(steps.sum.date.imp)             # mean total # of steps taken each day
medi <- median(steps.sum.date.imp)           # median total # of steps taken each day   
df.imp <- data.frame(date=names(steps.sum.date.imp), totalsteps=steps.sum.date.imp)
summary(df.imp$totalsteps)
print(avgi); print(medi)
```

```{r newhistogramtotalsteps, echo=TRUE, fig.width=10, fig.align='center'}
g <- ggplot(data=df.imp, aes(x=totalsteps))
g + geom_histogram(binwidth=550, col="black", fill="grey") +
    geom_vline(xintercept=avgi, size=0.5, col="red") +
    geom_vline(xintercept=medi, size=0.5, col="blue") +
    labs(title = "Histogram of Total Steps taken per Day (using new Imputed Dataset)",
         x = "Number of Daily Steps")
```  

The **mean** value is $`r round(avgi, 0)`$ and the **median** value is $`r round(medi, 0)`$.  
*Note: In the histogram above, the mean is denoted by the red color vertical line and the median by the blue color vertical line.*  

The new values differ from the estimates from the first part of the assignment only slightly or not at all (within margin of error) as seen in these calculations:    
Difference between orginal and new **mean** value is: $`r avg` - `r round(avgi,0)` = `r avg - round(avgi,0)`$.  
Difference between orginal and new **median** value is: $`r med` - `r medi` = `r med - medi`$.  

We conclude the impact of imputing missing data on the estimates of the total daily number of steps has had little to no effect.  



## Are there differences in activity patterns between weekdays and weekends?
We use the dataset with the filled-in missing (NA) values for this part of the report.  

#### 1. Create a new factor variable in the dataset with two levels: "weekday" and "weekend".
```{r newfactordataset, echo=TRUE}
daytype <- vector()                         # vector for daytype (ie Monday thru Sunday)
for(i in 1:nrow(act.imp)) {
    if( weekdays(as.Date(act.imp[i,]$date)) %in% c("Saturday", "Sunday") ) {
        daytype[i] <- "weekend"
    }
    else {
        daytype[i] <- "weekday"
    }
}
act.imp$daytype <- as.factor(daytype)       # new factor daytype added in dataset 
str(act.imp)
```  

#### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r timeseriespanel, echo=TRUE, fig.width=10}
steps.aggr <- aggregate(act.imp$steps, by=list(act.imp$interval, act.imp$daytype), mean)
names(steps.aggr) <- c("interval", "daytype", "avgsteps")
avgwd <- with( subset(steps.aggr, daytype == "weekday"), mean(avgsteps) )
avgwe <- with( subset(steps.aggr, daytype == "weekend"), mean(avgsteps) )
print(avgwd); print(avgwe)
g <- ggplot( data=steps.aggr, aes(x=interval, y=avgsteps) )
g + geom_line() +
    geom_hline(data=subset(steps.aggr, daytype == "weekday"), 
               aes(yintercept=avgwd), size=0.5, col="blue") +
    geom_hline(data=subset(steps.aggr, daytype == "weekend"), 
               aes(yintercept=avgwe), size=0.5, col="blue") +
    facet_grid(daytype ~ .) +
    labs(title = "Time Series of 5-Minute Interval and Average Steps by Weekday & Weekend",
         x = "5-min Interval",
         y = "Avgerage Number of Steps")
```  

The **mean** average number of steps for weekdays is $`r round(avgwd, 0)`$.  
The **mean** average number of steps for weekends is $`r round(avgwe, 0)`$.  
*Note: In plot above, means are denoted respectivley by blue color horizontal lines.*    
From the plot, we can see that on weekends the avg number of steps from 10:00 to 20:00 is increases compared to that same time interval on weekdays. On weekdays, the peak in avg steps occurs in early mornings (6:00 to 9:00) and a smaller peak in late afternoon (1800 to 1900). On weekends the avg steps are more consistent between the time of usual activity.  

